---
title: "RHRVEasy tutorial"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

RHRVEasy automates all steps of a Heart Rate Variability (HRV) analysis, including data processing, indices calculation, and statistical analysis. It takes as input a list of folders, each containing the recordings of the same population. It calculates time, frequency, and nonlinear domain HRV indices, and then it applies hypothesis test, corrects the significance levels. If there are more than two experimental groups and statistically significant differences, it performs a post hoc analysis to find out which groups have significant differences. 

# 0. Set up required to run this tutorial
This tutorial uses the recordings of the [Normal Sinus Rhythm RR Interval Database](https://physionet.org/content/nsr2db/1.0.0/) (hereinafter referred to as NSR_DB) and of the [Congestive Heart Failure RR Interval Database](https://archive.physionet.org/physiobank/database/chf2db/) (hereinafter referred to as CHF_DB). The first is made of healthy subjects, and the second of patients with a severe cardiac pathology. Hence significant differences in many HRV indices are expected between them.

To validate the performance of the package when working with more than two experimental groups, half of the recordings from both databases were randomly selected. The NSR_HALF_DB database is made up of 24 recordings randomly chosen from the 58 of the NSR database. The CHF_HALF_DB database is made up of 14 recordings chosen from the 28 in the CHF database. The four databases are available in the [RHRVEasy GitHub repository](https://github.com/sofiabardaji/RHRVEasy). To run this tutorial, the following four variables must contain the folder on the local machine where each of these registry databases are placed:
```{r}
NSR_HDB = "C:\\rrs\\RHRVEasy\\rrs\\normal"
CHF_DB = "C:\\rrs\\RHRVEasy\\rrs\\chf"
NSR_HALF_DB = "C:\\rrs\\RHRVEasy\\rrs\\normal_half"
CHF_HALF_DB = "C:\\rrs\\RHRVEasy\\rrs\\chf_half"
```

RHRVEasy permits creating an Excel spreadsheet with all the HRV indices calculated for each recording. The following variable must contain the folder on the local machine where the  Excel spreadsheet is to be placed:
```{r}
spreadsheet_path = "C:\\rrs\\RHRVEasy"
```

# 1. Time and frequency analysis

RHRVEasy enables the user to carry out a full HRV analysis by just invoking a function with a single mandatory parameter: a list with the folders containing the recordings of the experimental groups. This list must have least two folders. Each folder must contain all the RR recordings of the same experimental group and no additional files, as RHRVEasy will try to open all the files in those folders. The name that will be used to refer to each experimental group within the tool will be the name of the folder in which its recordings are located.

The following function call computes the time and frequency indices for the NSR and CHF databases, and and performs a statistical comparison of each index with a significance level correction using the Bonferroni method:

```{r}
source("C:\\rrs\\RHRVEasy\\RHRVEasy.r")
RHRVEasy_obj = RHRVEasy(folders =c(NSR_HDB, CHF_DB))
```

When the returned object is displayed in the console, it shows which indices present statistically significant differences:

```{r}
RHRVEasy_obj
```
All computed indices, as well as all pvalues resulting from all comparisons, are stored in dataframes contained in the object. Two different sets of pvalues are different: before (`pValues`) and after (`uncorrectedPvalues`) applying the significance level correction:

```{r}
head(RHRVEasy_obj$TimeAnalysis)
head(RHRVEasy_obj$FrequencyAnalysis)
# non-linear indices are set up NA since they have not been calculated
head(RHRVEasy_obj$pValues)
head(RHRVEasy_obj$uncorrectedPvalues)
```

By default, the frequency analysis is performed using the Fourier transform. It is also possible to use the wavelet transform pasing the value `wavelet` to the `typeAnalysis` parameter:
```{r}
RHRVEasy_obj_wavelet = RHRVEasy(folders =c(NSR_HDB, CHF_DB),  typeAnalysis = 'wavelet')
RHRVEasy_obj_wavelet
```

# 2. Correction of the significance level

Given that multiple statistical tests are performed on several HRV indices, a correction of the significance level should be applied. The Bonferroni method is used by default. This behavior can be overridden with the parameter `correctionMethod`. The possible values of this parameter besides `bonferroni` are `holm`, `hochberg`, `hommel`, `BH` (Benjamini & Hochberg), `fdr` (false discovery rate), `BY` (Benjamini & Yekutieli), and `none`. The boolean parameter `correction` can also be used to indicate that no correction is to be made; if the value of this parameter is TRUE be the Bonferroni method will be used.

```{r}
RHRVEasy_obj_no_correction = RHRVEasy(folders = c(NSR_HDB, CHF_DB), correction = FALSE)
RHRVEasy_obj_fdr = RHRVEasy(folders = c(NSR_HDB, CHF_DB), correctionMethod = 'fdr')
head(cbind(RHRVEasy_obj_no_correction$pValues, RHRVEasy_obj_fdr$pValues, RHRVEasy_obj$pValues))
sum(RHRVEasy_obj_no_correction$pValues < 0.05, na.rm=TRUE)
sum(RHRVEasy_obj_fdr$pValues < 0.05, na.rm=TRUE)
sum(RHRVEasy_obj$pValues < 0.05, na.rm=TRUE)
```

The level from which a difference in two indices is considered significant can be changed with the parameter `significance_level`:

```{r}
RHRVEasy_obj_001 = RHRVEasy(folders = c(NSR_HDB, CHF_DB), significance_level = 0.01)
RHRVEasy_obj_001
```

# 3. Correction of the significance level

If the argument `saveHRVindexesInPath` ss specified when invoking the function `RHRVEasy`, an Excel spreadsheet with all the HRV indices calculated for each recording will be created in the specified path. The name of the spreadsheet generated "Excel sheet is "<group 1 name> vs <group 2 name< .xlsx":

```{r}
RHRVEasy_obj = RHRVEasy(folders =c(NSR_HDB, CHF_DB), saveHRVindexesInPath = spreadsheet_path)
```


This spreadsheet can be generated from the object returned by `RHRVEasy` by calling the function `saveHRVindexes`.

```{r}
saveHRVindexes (RHRVEasy_obj, saveHRVindexesInPath = spreadsheet_path)
```

# 4. Correction of the significance level

If the analysis involves three or more groups, when statistically significant differences are found among them it does not necessarily means that there are statistically significant differences between all pairs of groups. In such a scenario the Dunn post-hoc is used to find which pairs of groups present differences:

```{r}
#comparison of the four databases
RHRVEasy_obj_four_groups = RHRVEasy(folders =c(NSR_HDB, CHF_DB, NSR_HALF_DB, CHF_HALF_DB))
RHRVEasy_obj_four_groups
#let's save the indices
saveHRVindexes (RHRVEasy_obj_four_groups, saveHRVindexesInPath = spreadsheet_path)
```

